A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-07-16 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: reworked introduction to non-additive integrals

@@

\textbf{Use of non-additive measure theory.} Given that extensive literature already exists for non-additive measure theory, we originally thought it would be just a matter of clarifying our definitions and connect them to the existing literature. However, things turned out to be more complicated, and we are now convinced that the tools already available are not applicable to our case.

In this section we will first clarify the requirements and what we expect to have from a non-additive generalization of integration and derivation. Then we will show that the tools we have found in the literature, like the Sugeno and Choquet integral, fail for simple examples.

\textbf{Requirements.} We have shown how fraction capacities and the state capacities give us generalized version of probability measure and state counting measures. The question now becomes: is there a generalization of all other elements of probability theory (e.g. probability densities, random variables, expectation) to the non-additive case? Since integral and differential calculus in probability theory is based on measure theory, is the a non-additive version of calculus that would allow such a generalization?

For simplicity, we will assume that Choquet theory applies: the ensemble space $\Ens$ has a clear set of extreme points $X$ and the capacities have a unique restriction to the extreme points. In principle, the general problem is more complicated since, as we mentioned before, $\Ens$ may not have extreme points. Whether one can always reconstruct the points and the topology of $X$ given $\Ens$ is an open question that we leave for our larger work.\footnote{One attempt, for example, was by using orthogonality to create the lattice of orthogonal subspaces and use something along the lines of Stone's representation theorem for Boolean algebra. However, this fails to recover the points correctly on the continuum even for classical mechanics. We are still trying to understand what mathematical structures have enough information to allow the reconstruction.} For our purposes, we can assume that both the state capacity and the fraction capacity as non-additive measure over $X$.

We goal is to define an expectation for statistical variables that reduces to the standard case in classical theories, that is, in the additive case. We already have a notion of statistical variable $F : \Ens \to \mathbb{R}$ defined over the ensemble space. This too can be restricted on the extreme points $X$ and, given that $F$ is a linear functional, the values on the extreme points will fully determine the values for all other points of $\Ens$. In the classical case, the expectation is recovered through the integration of the variable over the probability density. Therefore, we would like a non-additive integral such that
\begin{equation}
	E[F] = F(\ens) = (E)\int_X F(x) d\frcap_{\ens}.
\end{equation}
which reduces to the Riemann integral in the classical (additive) case. The integral will depend only on the values of $F$ and $\frcap_{\ens}$ on $X$, but since these uniquely determine both $F$ and $\ens$, the integral can be understood as a map from $F$ and $\ens$ to $F(\ens)$, which is, in principle, well-defined. Therefore, in principle, this integral should exist.

Additionally, we would like to be able to define an analogue of the probability density. We are therefore looking for a non-additive derivative $\rho = \frac{d \frcap_{\ens}}{d \capacity}$ that reduces to the Radon-Nikodym derivative in the classical (additive) case such that
\begin{equation}
	F(\ens) = (E)\int_X F(x) \rho(x) d\capacity
\end{equation}
under the previous integral. It is less clear whether this derivative is always well defined. Since in classical mechanics we have $S(\ens) = \int_X \rho \log \rho d\mu$ and in quantum mechanics $S(\ens) = \tr[\rho \log \rho]$, we would expect
\begin{equation}
	S(\ens) = (E)\int_X \rho(x) \log \rho(x) d\capacity.
\end{equation}
We do not know whether this second requirement is at all possible, or whether it can be used as guide in case that the previous derivative is not uniquely defined.

\textbf{Ruling out Sugeno and Choquet integrals.} Non-additive measures are used in disparate contexts, such as game theory, economic decision theory, optimization and artificial intelligence. CITE To our knowledge, there isn't a single unified theory, though some work has been proceeding in that direction.\cite{nonadditive2014,pap2013} For now, we will concentrate on integration. The Choquet integral and the Sugeno integral are the oldest and most common approaches to integration of non-additive measure, and therefore we should first test whether these work.

Note that the Sugeno integral does not recover the Lebesgue integral in the case the capacity is additive (i.e. classical), and is therefore ruled out. The Choquet integral does, so it will work in the classical case. However, it does not work in the quantum case. We can show this by computing the Choquet integral for a two states quantum system and see that it does not recover the expectation of the statistical variables in the corresponding ensemble space.

@@

>>>>

\textbf{Expectation and integration.} TODO: show the use of integrals over probability measure.

@@

\textbf{Requirements.} In a previous section we reviewed what types of objects we have in probability theory for classical ensembles: a sample space $X$, a probability measure $p$, random variables $F : X \to \mathbb{R}$, expectations as integrals of the probability measure $E[F] = \int_X F(x) dp$, a state counting measure $\mu$, the probability density as the Radon-Nikodym derivative $\frac{dp}{d\mu}$ and the entropy as $\int_X \rho \log \rho d\mu$.  Given that the fraction capacity $\frcap_{\ens}$ can be seen as a generalization of probability measure $p$ and the state capacity $\capacity$ can be seen as a generalization of the states counting measure $\mu$, is there a generalization of all other elements? That is, can we find an integral and derivative operation over non-additive measures that generalizes to all physical theories?

The first difficulty, as we already pointed out, is that the ensemble space $\Ens$ may not have extreme points. What is the correct procedure to ``close'' the ensembles space and recover the space of extreme points is, for us, an open question. \footnote{One attempt, for example, was by using orthogonality to create the lattice of orthogonal subspaces and use something along the lines of Stone's representation theorem for Boolean algebra. However, this fails to recover the points correctly on the continuum even for classical mechanics. We are still trying to understand what mathematical structures have enough information to allow the reconstruction.} For the purpose of this paper, we will assume $\Ens$ is compact, it has a set of extreme points $X$ and the capacities have a unique restriction to the extreme points. Therefore we can talk about the restriction on $X$ of the state capacity, the fraction capacity and the statistical variables.

Recall that a statistical variable is linear function $F : \Ens \to \mathbb{R}$ defined over the ensemble space. Given that $F$ is a linear, its restriction on $X$ will determine the values for all other points of $\Ens$. Since the expectation is recovered through the integration of the variable over the probability density, we would like a non-additive integral $(E)\int_X$ such that
\begin{equation}
	E[F] = F(\ens) = (E)\int_X F(x) d\frcap_{\ens}.
\end{equation}
which reduces to the Riemann integral in the classical (additive) case. The integral will depend only on the values of $F$ and $\frcap_{\ens}$ on $X$, but since these uniquely determine both $F$ and $\ens$, the integral can be understood as a map from $F$ and $\ens$ to $F(\ens)$, which is, in principle, well-defined. Therefore, in principle, this integral should exist.

Additionally, we would like to be able to define an analogue of the probability density. We are therefore looking for a non-additive derivative $\rho = \frac{d \frcap_{\ens}}{d \capacity}$ that reduces to the Radon-Nikodym derivative in the classical (additive) case such that
\begin{equation}
	F(\ens) = (E)\int_X F(x) \rho(x) d\capacity
\end{equation}
under the previous integral. It is less clear whether this derivative is always well defined. Ideally, we would also expect
\begin{equation}
	S(\ens) = (E)\int_X \rho(x) \log \rho(x) d\capacity.
\end{equation}
This last requirement may not be possible, or it may be used as guide in case that the previous derivative is not uniquely defined.

\textbf{Ruling out Sugeno and Choquet integrals.} Given that extensive literature already exists for non-additive measure theory (cite), as it is used in disparate contexts, such as game theory, economic decision theory, optimization and artificial intelligence, we originally thought it would be just a matter of clarifying our definitions and connect them to the existing literature. However, things turned out to be more complicated. Depending on the use cases, different types of integrals exists. In fact, procedures exist\cite{nonadditive2014,pap2013} to construct infinitely many types of non-additive integrals.\footnote{Briefly, one can define a pseudo-addition and a pseudo-multiplication as real functions of two real arguments and construct an integral based on those definitions.} We can, however, at least rule out the two most common non-additive integrals: the Sugeno integral and the Choquet integral.

The Sugeno integral does not recover the Lebesgue integral in the case the capacity is additive (i.e. classical), and is therefore quickly ruled out. The Choquet integral does, so it will work in the classical case. However, it does not work in the quantum case. We can show this by computing the Choquet integral for a two states quantum system and see that it does not recover the expectation of the statistical variables in the corresponding ensemble space.

@@

\textbf{Ruling out Sugeno and Choquet integrals.}
